Grade Retention and Mixed Race

### 2022-11-17 products/paper/main.qmd
Add first full rough draft of conclusion

-   Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups. This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization.

>>>>

In this article, I have used the risk of grade retention among K-12 students to better understand how biracial students' life chances compare to their monoracial peers. Unlike prior work on this topic, due to the large samples involved, the estimates for most biracial groups are relatively precise. For groups with a high level of precision, the results tell a consistent story - biracial students' risk of grade retention is similar to their monoracial constituent group with the lower risk.

These results contradict expectations of both a "halfway" position and the dominance of the one drop rule. For no group do I observe clear evidence of a pattern of hypodescent in which the risk was similar to the higher risk constituent monoracial group. These results also strongly contradict the expectations of the marginal man hypothesis.

Two possibilities emerge that may help us understand these results better. First, we may be witnessing a shift toward a new regime of hyperdescent that applies broadly across a wide variety of racial groups. Second, the relatively low risk for biracial students may be a product of the strength of hybridity in allowing biracial students to better negotiate racialized systems than their monoracial minority peers.

Differences in family resources were expected to play some role in the relative position of biracial students' risk. The results on parent selection vary somewhat by group, but the majority of biracial groups had resources more similar to their monoracial constituent group with higher resources, indicating a strong positive selection into interracial unions in the prior generation. The most important exception to this trend is the case of Black/White biracial students who have far fewer resources than white students and resources closer to black students.

Regardless of variation in biracial students' parental resources, accounting for these resources differentials does not substantially change the overall result that biracial students' risk is more similar to the constituent monoracial group with lower risk. The one exception to this finding is for the Latino/Asian case where the much lower risk of these students compared to Latino and Asian students was entirely driven by resource differentials.

The relative risk for biracial individuals is sometimes incorrectly treated as an indication of a positive future direction for the US in terms of ameliorating racial inequality. The improved prospects for biracial individuals do nothing to ameliorate the often strong and persistent inequalities between monoracial groups. The growth and relative success of biracial populations may isolate remaining members of monoracial minority groups as much as it assimilates mixed race individuals. The real question at stake is how existing divides and identities may be restructured in the context of a growing biracial population. This research suggests the potential for a growing divide between these mixed race populations and the monoracial minority populations from which they derive at least part of their ancestry.

Because these findings rely upon large samples, they present an important step forward in our understanding of how biracial individuals will fit into America's system of racial inequality. However, this study also has important limitations that need to be noted.

Although sample sizes are big generally, they remain quite small for some groups of non-white biracial respondents, limiting our understanding of the outcomes for those groups. In particular, results for non-white part-Indigenous populations are quite imprecise. The results interestingly and counterintuitively suggest that part-Indigenous populations are the least likely to be in the position expected by hyperdescent, but due to low sample sizes, that finding is highly tentative.

Wile grade retention is an important early life outcome, it is far from the only one that these individuals will face in their life. Although, we often observe similar racial inequalities across a variety of outcomes, knowing the result for one outcome does not perfectly predict other outcomes. As biracial individuals age and face later outcomes in the educational system, the labor market, and elsewhere, their experiences may differ from what I observe here. Thus, collecting more detailed information on these outcomes in future studies remains important.
